Final Project

### 2025-05-01 MSSC5750_Project.qmd
Data Cleaning, One-Hot-Encoding, Results Table

table(data$previous_loan_defaults_on_file)

@@

@@

# Preparing categorical variables. Should happen automatically for factors
data$person_gender <- as.factor(data$person_gender)
data$person_home_ownership<-as.factor(data$person_home_ownership)
data$loan_intent<-as.factor(data$loan_intent)
data$previous_loan_defaults_on_file<-as.factor(data$previous_loan_defaults_on_file)

# Extracting outcome variable
class <- data$loan_status

# Converting to a matrix
matrixData <- model.matrix(class ~ ., data = data %>% select(-loan_status))

@@

# See coefficients, which variables we might drop
coef(lasso)

# Make predictions
data$lassoPred <- as.vector(predict(lasso, newx = matrixData, type = "class"))

# Evaluate accuracy measures
confusionMatrix(factor(data$lassoPred), factor(data$loan_status), positive = '1')

@@

# We need a train/test split. We will use 80/20
# can use uniform distribution, set seed for reproducability
set.seed(100)
forTrain<-runif(nrow(data))<.8

trainData<-data[forTrain,]
testData<-data[!forTrain,]

# First creating logistic regression with all our variables
logModel <- glm(loan_status ~ person_age + person_gender + person_education + person_income + person_emp_exp + person_home_ownership +loan_amnt + loan_intent + loan_int_rate + loan_percent_income + cb_person_cred_hist_length + credit_score + previous_loan_defaults_on_file, data=trainData, family= binomial )

# Repeating but removing the variables not relevant in lasso
# removed(person cred length, person_emp_exp person_income person_gender person_education person_age)
logModelLasso <- glm(loan_status ~ person_home_ownership +loan_amnt + loan_intent + loan_int_rate + loan_percent_income  + credit_score + previous_loan_defaults_on_file, data=trainData, family= binomial )

@@

@@

plot(explainer, plot_phi0 = FALSE, index_x_explain = 1)

shapObj <-shapviz(explainer)
sv_importance(shapObj, kind="beeswarm", show_numbers = TRUE, max_display=15)

>>>>

# Hm. In every single case where there was a prior loan default, loan_status=0. We probably don't want to include this variable in anything explanatory
table(data$previous_loan_defaults_on_file, data$loan_status)

@@

Data cleaning
```{R}

data2<-data
######## Transformation of variables ########

#### person_income ####

# Salary is extremely right skewed, lets use the log transformation. Minimum is 8000 so we do not need to add a constant
min(data2$person_income)
data2$log_person_income <- log(data2$person_income)

# Our max is still way larger than our 3rd quartile, so maybe we should cap salary before transformation
summary(data2$log_person_income)
data2$log_person_income<-log(ifelse(data2$person_income>500000,500000,data2$person_income))

# Better
summary(data2$log_person_income)

#### person_emp_exp ####

# Employee experience also very skewed. And some of the values dont really make any sense. Log transformation can fix this too
summary(data2$person_emp_exp)

# First replace extreme (impossible) values with something realistic
data2$person_emp_exp<-ifelse(data2$person_emp_exp>50,50,data2$person_emp_exp)

data2$log_person_emp_exp <- log(data2$person_emp_exp+.5)

# Better
summary(data2$log_person_emp_exp)

#### Age  ####
# We need to bucket our age. Theres no way we actually have someone aged 144, and 75% of our data falls between 20-30
data2<- data2 %>% mutate(person_age = case_when(person_age<25 ~ "<25",
                                                person_age<=30 ~"25-30",
                                                person_age<45 ~ "31-44",
                                                person_age<60 ~ "45-59",
                                                TRUE ~ "60+"))

data2$person_age<-as.factor(data2$person_age)
table(data2$person_age)

####  cred hist length  ####
# Only a handful fall >=18, we will just cap it all there as a "catch all", but leave as numeric
table(data2$cb_person_cred_hist_length)
data2$cb_person_cred_hist_length <- ifelse(data2$cb_person_cred_hist_length>18,18, data2$cb_person_cred_hist_length)

######## One hot encoding for categorical variables ########

# Preparing categorical variables.
data2$person_gender <- as.factor(data2$person_gender)
data2$person_home_ownership<-as.factor(data2$person_home_ownership)
data2$loan_intent<-as.factor(data2$loan_intent)
data2$previous_loan_defaults_on_file<-as.factor(data2$previous_loan_defaults_on_file)

# One hot encoding for categorical variables
data3 <- dummy_cols(data2, remove_first_dummy = FALSE, remove_selected_columns = TRUE)

# we should remove first dummy to prevent multicolinnearity , need to keep track of what was dropped because that is our reference value
data3<-data3 %>% select(-c("person_gender_female","person_education_High School","person_home_ownership_MORTGAGE", "loan_intent_DEBTCONSOLIDATION", "previous_loan_defaults_on_file_No", "person_age_<25"))

# replacing base data with cleaner version now
cleanData <- data3

@@

# Extracting outcome variable
class <- cleanData$loan_status

# Converting to a matrix, exclude the non-log-transformed, and also the prior loan default
matrixData <- model.matrix(class ~ ., data = cleanData %>% select(-c("loan_status", "person_income", "person_emp_exp", "previous_loan_defaults_on_file_Yes") ))

@@

# See coefficients, which variables we might drop
# Now that we switched to log income, variable is no longer being dropped. We will proceed with log_income
coef(lasso)

# Make predictions
cleanData$lassoPred <- as.vector(predict(lasso, newx = matrixData, type = "class"))

# Evaluate accuracy measures
confusionMatrix(factor(cleanData$lassoPred), factor(cleanData$loan_status), positive = '1')

# And again with our loan_default knowledge:
confusionMatrix(factor( ifelse(cleanData$previous_loan_defaults_on_file_Yes==1,0,cleanData$lassoPred)), factor(cleanData$loan_status), positive = '1')

@@

# We need a train/test split. We will use 80/20
# can use uniform distribution, set seed for reproducability
set.seed(100)
forTrain<-runif(nrow(cleanData))<.8

trainData<-cleanData[forTrain,]
testData<-cleanData[!forTrain,]

# First creating logistic regression with all our variables
logModel <- glm(loan_status ~ . , data=trainData %>% select(-c("lassoPred", "person_income", "person_emp_exp", "previous_loan_defaults_on_file_Yes")), family= binomial )

# Repeating but only with variables selected by lasso
logModelLasso <- glm(loan_status ~ loan_int_rate + loan_percent_income + log_person_income + person_home_ownership_OTHER + person_home_ownership_OWN + person_home_ownership_RENT + loan_intent_EDUCATION + loan_intent_HOMEIMPROVEMENT + loan_intent_PERSONAL + loan_intent_VENTURE, data=trainData, family= binomial )

@@

# Our training data is still heavily imbalanced, repeating again for rebalanced data
table(trainData$loan_status)
ones<-trainData %>% filter(loan_status==1)
set.seed(5)
zeros<-trainData %>% filter(loan_status==0) %>% sample_n(8000)
trainDataBalanced<-rbind(ones,zeros)
table(trainDataBalanced$loan_status)

rebalancedLog <- glm(loan_status ~ . , data=trainDataBalanced %>% select(-c("lassoPred", "person_income", "person_emp_exp", "previous_loan_defaults_on_file_Yes")), family= binomial )

testData$rebalancedLog <- round(predict(rebalancedLog, testData, type='response'))

# Matrix for regular log
confusionMatrix(factor(testData$rebalancedLog), factor(testData$loan_status), positive = '1')

Creating dataframe of results
```{R}

results<-data.frame(model=character(),accuracy=numeric(),sensitivity=numeric(),specificity=numeric())

addresults<-function(name,res){
  results<<-rbind(results, data.frame(model=name,accuracy=res$overall['Accuracy'],sensitivity=res$byClass['Sensitivity'],specificity=res$byClass['Specificity']) )
}

addresults('Lasso', confusionMatrix(factor( cleanData$lassoPred), factor(cleanData$loan_status), positive = '1'))
addresults('Log', confusionMatrix(factor(testData$logisticPred), factor(testData$loan_status), positive = '1'))
addresults('LogLasso', confusionMatrix(factor(testData$logisticLassoPred), factor(testData$loan_status), positive = '1'))
addresults('LogBalanced', confusionMatrix(factor(testData$rebalancedLog), factor(testData$loan_status), positive = '1'))
rownames(results)<-NULL
results

###################### Repeating for accuracy after applying logic for previous default ##############################

results2<-data.frame(model=character(),accuracy=numeric(),sensitivity=numeric(),specificity=numeric())

addresults2<-function(name,res){
  results2<<-rbind(results2, data.frame(model=name,accuracy=res$overall['Accuracy'],sensitivity=res$byClass['Sensitivity'],specificity=res$byClass['Specificity']) )
}

addresults2('Lasso', confusionMatrix(factor(ifelse(cleanData$previous_loan_defaults_on_file_Yes==1,0,cleanData$lassoPred) ), factor(cleanData$loan_status), positive = '1'))
addresults2('Log', confusionMatrix(factor( ifelse(testData$previous_loan_defaults_on_file_Yes==1,0, testData$logisticPred) ), factor(testData$loan_status), positive = '1'))
addresults2('LogLasso', confusionMatrix(factor(ifelse(testData$previous_loan_defaults_on_file_Yes==1,0,testData$logisticLassoPred) ), factor(testData$loan_status), positive = '1'))
addresults2('LogBalanced', confusionMatrix(factor(ifelse(testData$previous_loan_defaults_on_file_Yes==1,0,testData$rebalancedLog) ), factor(testData$loan_status), positive = '1'))
rownames(results2)<-NULL
results2

@@

plot(explainer, plot_phi0 = FALSE, index_x_explain = 1)
####

###
shapObj <-shapviz(explainer)
sv_importance(shapObj, kind="beeswarm", show_numbers = TRUE, max_display=15)
